---
title: "Mapping Population Densities"
author: "TJ Palanca"
date: "1 May 2017"
output:
  html_document:
    toc: true
    toc_float: true
    theme: "lumen"
    highlight: "tango"
    code_folding: show
    self_contained: false
    lib_dir: libs
---

## Preliminaries

### Libraries

```{r setup}
# Package loads
suppressPackageStartupMessages({
  library(tidyverse)
  library(stringr)
  library(raster)
  library(dplyr)
  library(sf)
})

# Working directory setup
if (!str_detect(getwd(), "src")) setwd("src")
```

### Data Import

Data is downloaded from NASA SEDAC [here](http://sedac.ciesin.columbia.edu/data/set/gpw-v4-population-density-adjusted-to-2015-unwpp-country-totals/data-download).

```{r data-import}

# Unzip NASA SEDAC file (must download separately)
if (!dir.exists("../dta/gpw-v4/")) {
  unzip(
    zipfile = "../dta/gpw-v4-population-density-adjusted-to-2015-unwpp-country-totals-2015.zip",
    exdir = "../dta/gpw-v4/"
  )
}

# Download GADM Province data for Philippines
if (!file.exists("../dta/PHL_adm1.rds")) {
  download.file(
    url = "http://biogeo.ucdavis.edu/data/gadm2.8/rds/PHL_adm1.rds",
    destfile = "../dta/PHL_adm1.rds"
  )
}

```

### Data Processing

Read in the data into memory and process them into the final formats.

```{r data-loading-processing}

# Philippines administrative boundaries
readRDS("../dta/PHL_adm1.rds") ->
  philippines.spdf
  
# Read into memory (cropped to Philippines)
raster("../dta/gpw-v4/gpw-v4-population-density-adjusted-to-2015-unwpp-country-totals_2015.tif") %>% 
  crop(philippines.spdf) %>% 
  rasterToPoints(spatial = TRUE) %>% 
  intersect(philippines.spdf) %>% 
  setNames(c("popdensity_km", names(.)[2:length(names(.))])) ->
  philippines_popd.spdf

# Fortify to data frame
philippines_popd.spdf %>% 
  st_as_sf() %>% 
  mutate(lng = map_dbl(geometry, ~.[[1]]),
         lat = map_dbl(geometry, ~.[[2]])) %>% 
  as_data_frame() ->
  philippines_popd.df

```

## Plotting

We set up a general plotting function that:

  * considers lat/lng resolution, because for larger areas, the current grid resolution may be too high.
  * automatically decides the scaling of population density.

```{r}

plotPopulationDensityLines <- 
  function(df, lat_res, lng_res = lat_res, height = 5, lwd = 0.5, ...) {
    # Plot the population density lines
    #
    # Arguments:
    #   df:       data frame containing columns 
    #             lat, lng, and popdensity_km
    #   lat_res:  latitude resolution (to the nearest lat_res)
    #   lng_res:  longitude resolution (to the nearest lng_res)
    #   height:   maximum height (as factor of lat_res)
    #   lwd:      line width of line
    #   ...:      labels
    #
    # Effect:
    #   Plots the plot onto the relevant graphics device
  
    df %>% 
      # Perform resolution changes
      group_by(
        lat = round(lat * (1/lat_res))/(1/lat_res), 
        lng = round(lng * (1/lng_res))/(1/lng_res)
      ) %>% 
      summarise(popdensity_km = mean(popdensity_km, na.rm = T)) %>% 
      ungroup() %>% 
      # Scale population density 
      mutate(popdensity_km = popdensity_km/max(popdensity_km, na.rm = T)) %>% 
      # Assign groups to split up islands
      arrange(lat, lng) %>% 
      mutate(group = cumsum(
        if_else(lat != lag(lat) | 
                  lng - lag(lng) > lng_res * 1.00001 | 
                  is.na(lag(lng)), 1L, 0L))) %>% 
      ggplot(aes(x = lng, group = group,
                 y = lat + lat_res * height * popdensity_km)) +
      geom_path(lwd = lwd) +
      coord_map() +
      theme_bw(base_family = "Roboto") +
      theme(
        axis.title    = element_blank(),
        axis.ticks    = element_blank(),
        axis.text     = element_blank(),
        plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        panel.grid    = element_blank(),
        panel.border  = element_blank(),
        plot.caption  = element_text(size = 6, face = "italic", hjust = 0)
      ) +
      labs(
        ...,
        caption = "
DATA SOURCE:NASA Socioeconomic Data and Applications Center (SEDAC)
NOTE: Height represents the mean UN-adjusted population density

TROY JAMES PALANCA | TJPALANCA.COM
"
      )
  }
```

### Philippines

```{r}

philippines_popd.df %>% 
  plotPopulationDensityLines(
    lat_res  = 0.1,
    lng_res  = 0.05,
    height   = 50,
    lwd      = 0.3,
    title    = "POPULATION DENSITY",
    subtitle = "Philippines"
  ) -> philippines.gg

ggsave(
  plot = philippines.gg, dpi = 300,
  filename = "../out/01-philippines.png"
)

philippines.gg

```

### Greater Manila Area

```{r}

philippines_popd.df %>% 
  filter(NAME_1 %in% c(
    "Metropolitan Manila", 
    "Laguna", "Cavite", "Rizal", "Bulacan"
  )) %>% 
  plotPopulationDensityLines(
    lat_res  = 0.008,
    lng_res  = 0.01,
    height   = 20,
    lwd      = 0.3,
    title    = "POPULATION DENSITY",
    subtitle = "Greater Manila Area"
  ) -> gma.gg

ggsave(
  plot = gma.gg, dpi = 300,
  filename = "../out/02-greater-manila-area.png"
)

gma.gg

```

### Cebu

```{r}

philippines_popd.df %>% 
  filter(NAME_1 == "Cebu") %>% 
  plotPopulationDensityLines(
    lat_res  = 0.01,
    lng_res  = 0.01,
    height   = 20,
    lwd      = 0.3,
    title    = "POPULATION DENSITY",
    subtitle = "Cebu"
  ) -> cebu.gg

ggsave(
  plot = cebu.gg, dpi = 300,
  filename = "../out/03-cebu.png"
)

cebu.gg

```

### Davao

```{r}

philippines_popd.df %>% 
  filter(NAME_1 == "Davao del Sur") %>% 
  plotPopulationDensityLines(
    lat_res  = 0.01,
    lng_res  = 0.01,
    height   = 50,
    lwd      = 0.3,
    title    = "POPULATION DENSITY",
    subtitle = "Davao"
  ) -> davao.gg

ggsave(
  plot = davao.gg, dpi = 300,
  filename = "../out/04-davao.png"
)

davao.gg

```

### Bicol

```{r}

philippines_popd.df %>% 
  filter(NAME_1 %in% c(
    "Albay", "Camarines Norte", "Camarines Sur",
    "Masbate", "Sorsogon", "Catanduanes"
  )) %>% 
  plotPopulationDensityLines(
    lat_res  = 0.01,
    lng_res  = 0.01,
    height   = 65,
    lwd      = 0.2,
    title    = "POPULATION DENSITY",
    subtitle = "Bicol Region"
  ) -> bicol.gg

ggsave(
  plot = bicol.gg, dpi = 300,
  filename = "../out/05-bicol.png"
)

bicol.gg

```